---
title: "Reading and Writing NetCDF-CF Geometry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading and Writing Geometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6, 
  fig.height=4
)
options(scipen = 9999)
```

## Introduction

This example shows how to 

1. add polygon data to a NetCDF-CF Discrete Sampling Geometry timeSeries featureType file, 
1. read the polygon data from the NetCDF file, 
1. and write it to a standard GIS format.  

Below the example, all the geometry types that `ncdfgeom` handles are shown in detail. 

## Load Spatial Data

Tables of point, line, or polygon features with associated timeseries are the target for this functionality. Here, we use some sample data from the `ncdfgeom` package.
```{r libs}
example_file <- "example.nc"

file.copy(from = system.file('extdata/example_huc_eta.nc', package = 'ncdfgeom'), 
					to = example_file, 
					overwrite = TRUE) -> quiet

polygons <- sf::read_sf(system.file('extdata/example_huc_eta.json', package = 'ncdfgeom'))

polygons <- dplyr::select(polygons, LOADDATE, AREASQKM, HUC12, NAME)

plot(sf::st_geometry(polygons))
```

Now we have the polygons as shown above and a NetCDF file with a header that looks like:
```{r dump_polygons, echo=FALSE}
ncdump <- system(paste("ncdump -h", example_file), intern = TRUE)
cat(ncdump ,sep = "\n")
```
Now we can use the `write_geometry` function to add the polygon data to the NetCDF file.
```{r demo}
(vars <- ncmeta::nc_vars(example_file))

ncdfgeom::write_geometry(nc_file=example_file,
                         geom_data = polygons, 
                         instance_dim_name = "station", 
                         variables = vars$name) -> example_file
```

Now the NetCDF file looks like:
```{r dump_polygons_ts, echo=FALSE}
ncdump <- system(paste("ncdump -h", example_file), intern = TRUE)
cat(ncdump ,sep = "\n")
```

Read the polygon data from the file and write it out to a geopackage.

```{r read, warning=F}
polygons_sf <- ncdfgeom::read_geometry(example_file)

plot(sf::st_geometry(polygons_sf))
sf::write_sf(polygons_sf, "polygons.gpkg")
```

```{r cleanup, echo=F}
temp <- file.remove(example_file, "polygons.gpkg")
```

##   Geometry Types

```{r examples, child="../tests/testthat/data/geom_examples.md"}
