---
title: "ncdfgeom geometry tools"
author: "David Blodgett"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ncdfgeom geometry tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6, 
  fig.height=4
)
options(scipen = 9999)
```

```{r load, echo=FALSE, message=FALSE, warning=FALSE}
file.copy(system.file('extdata','example_huc_eta.nc', package = 'ncdfgeom'), 
					"hucPolygons.nc", 
					overwrite = TRUE) -> quiet
```
Below is a demo that shows how to add polygon data to a NetCDF DSG TimeSeries file, read the polygon data from the NetCDF file, and write it to a shapefile format.

First, we load up a NetCDF file and some geoJson data containing polygons.
```{r libs}
hucTimeseries <- ncdf4::nc_open("hucPolygons.nc")

hucPolygons <- sf::read_sf(system.file('extdata','example_huc_eta.json',
                                       package = 'ncdfgeom'))
plot(sf::st_geometry(hucPolygons))
```

Now we have the polygons as shown above and a NetCDF file with a header that looks like:

```
netcdf hucPolygons {
dimensions:
	maxStrlen64 = 64 ;
	station = 2 ;
	time = 25 ;
variables:
	double lat(station) ;
		lat:units = "degrees_north" ;
		lat:missing_value = -999. ;
		lat:long_name = "latitude of the observation" ;
		lat:standard_name = "latitude" ;
	double lon(station) ;
		lon:units = "degrees_east" ;
		lon:missing_value = -999. ;
		lon:long_name = "longitude of the observation" ;
		lon:standard_name = "longitude" ;
	double time(time) ;
		time:units = "days since 1970-01-01 00:00:00" ;
		time:missing_value = -999. ;
		time:long_name = "time of measurement" ;
		time:standard_name = "time" ;
	char station_name(station, maxStrlen64) ;
		station_name:units = "" ;
		station_name:missing_value = "" ;
		station_name:long_name = "Station Names" ;
		station_name:cf_role = "timeseries_id" ;
		station_name:standard_name = "station_id" ;
	int et(station, time) ;
		et:units = "mm" ;
		et:missing_value = -999 ;
		et:long_name = "Area Weighted Mean Actual Evapotranspiration" ;
		et:coordinates = "time lat lon" ;

// global attributes:
		:Conventions = "CF-1.7" ;
		:featureType = "timeSeries" ;
		:cdm_data_type = "Station" ;
		:standard_name_vocabulary = "CF-1.7" ;
}
```

Now we can use the `write_geometry` function to add the polygon data to the NetCDF file.
```{r demo}
hucPolygons_nc <- ncdfgeom::ToNCDFSG(nc_file="hucPolygons.nc",
                                       geomData = hucPolygons, 
                                       instance_names = hucPolygons$HUC12, 
                                       instance_dim_name = "station", 
                                       variables = hucTimeseries$var)
```

Now the header of the NetCDF file looks like: (note some variables removed to be more concise)
```
netcdf hucPolygons {
dimensions:
	maxStrlen64 = 64 ;
	station = 2 ;
	time = 25 ;
	char = 38 ;
	node = 2628 ;
variables:
	double lat(station) ;
		lat:units = "degrees_north" ;
		lat:missing_value = -999. ;
		lat:long_name = "latitude of the observation" ;
		lat:standard_name = "latitude" ;
		lat:grid_mapping = "grid_mapping" ;
		lat:geometry = "geometry_container" ;
	double lon(station) ;
		lon:units = "degrees_east" ;
		lon:missing_value = -999. ;
		lon:long_name = "longitude of the observation" ;
		lon:standard_name = "longitude" ;
		lon:grid_mapping = "grid_mapping" ;
		lon:geometry = "geometry_container" ;
	double time(time) ;
		time:units = "days since 1970-01-01 00:00:00" ;
		time:missing_value = -999. ;
		time:long_name = "time of measurement" ;
		time:standard_name = "time" ;
	char station_name(station, maxStrlen64) ;
		station_name:units = "" ;
		station_name:missing_value = "" ;
		station_name:long_name = "Station Names" ;
		station_name:cf_role = "timeseries_id" ;
		station_name:standard_name = "station_id" ;
		station_name:grid_mapping = "grid_mapping" ;
		station_name:geometry = "geometry_container" ;
	int et(station, time) ;
		et:units = "mm" ;
		et:missing_value = -999 ;
		et:long_name = "Area Weighted Mean Actual Evapotranspiration" ;
		et:coordinates = "time lat lon" ;
		et:grid_mapping = "grid_mapping" ;
		et:geometry = "geometry_container" ;
	double AREASQKM(station) ;
		AREASQKM:units = "unknown" ;
		AREASQKM:_FillValue = NaN ;
		AREASQKM:grid_mapping = "grid_mapping" ;
		AREASQKM:geometry = "geometry_container" ;
	char STATES(station, char) ;
		STATES:units = "unknown" ;
		STATES:grid_mapping = "grid_mapping" ;
		STATES:geometry = "geometry_container" ;
	char HUC12(station, char) ;
		HUC12:units = "unknown" ;
		HUC12:grid_mapping = "grid_mapping" ;
		HUC12:geometry = "geometry_container" ;
	char NAME(station, char) ;
		NAME:units = "unknown" ;
		NAME:grid_mapping = "grid_mapping" ;
		NAME:geometry = "geometry_container" ;
	double x(node) ;
		x:units = "degrees_east" ;
		x:standard_name = "longitude" ;
		x:cf_role = "geometry_x_node" ;
	double y(node) ;
		y:units = "degrees_north" ;
		y:standard_name = "latitude" ;
		y:cf_role = "geometry_y_node" ;
	float geometry_container ;
		geometry_container:geometry_type = "polygon" ;
		geometry_container:node_count = "node_count" ;
		geometry_container:node_coordinates = "x y" ;
		geometry_container:grid_mapping = "grid_mapping" ;
	int node_count(station) ;
		node_count:long_name = "count of coordinates in each instance geometry" ;
	float grid_mapping ;
		grid_mapping:grid_mapping_name = "latitude_longitude" ;
		grid_mapping:semi_major_axis = 6378137. ;
		grid_mapping:inverse_flattening = 298.257223563 ;
		grid_mapping:longitude_of_prime_meridian = 0. ;

// global attributes:
		:Conventions = "CF-1.8" ;
		:featureType = "timeSeries" ;
		:cdm_data_type = "Station" ;
		:standard_name_vocabulary = "CF-1.7" ;
}
```

Now we can read the polygon data from the file.

```{r read, warning=F}
hucPolygons_sf <- ncdfgeom::FromNCDFSG("hucPolygons.nc")

plot(sf::st_geometry(hucPolygons_sf))
hucPolygons_shp <- sf::write_sf(hucPolygons_sf, "hucPolygons.gpkg")
```

```{r cleanup, echo=F}
temp <- file.remove("hucPolygons.nc", "hucPolygons.gpkg")
```

## Geometry Types

`ncdfgeom` works with a set of geometry types. 

```{r examples, child="../tests/testthat/data/geom_examples.md"}
