---
title: "Area Weight Generation for Polygon Intersections"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Area Weight Generation for Polygon Intersections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_1, include = FALSE}
local <- (Sys.getenv("BUILD_VIGNETTES") == "TRUE")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=local,
  fig.width=6, 
  fig.height=4
)

library(ncdfgeom)

if(!require(nhdplusTools)) {
  warning("need nhdplusTools for this demo")
  knitr::opts_chunk$set(eval = FALSE)
}

org_options <- options(scipen = 9999)
```

This article demonstrates how to create area weights for two sets of polygons.

There are three examples included, one for a [typical application with hydrologic units](#applied-example),
one for [a set of polygons that have gaps](#extended-example-one) and one [for polygons without gaps](#extended-example-two).  

## Applied Example

[It is a comparison with the `gdptools` python package demonstration here.](https://gdptools.readthedocs.io/en/latest/Examples/PolyToPoly/Updated_PolytoPoly_weights.html)

```{r} 

gdptools_weights <- read.csv(system.file("extdata/gdptools_prl_out.csv", package = "ncdfgeom"), 
                             colClasses = c("character", "character", "numeric"))

gdptools_weights <- dplyr::rename(gdptools_weights, gdptools_wght = wght)

gage_id <- "USGS-01482100"
basin <- nhdplusTools::get_nldi_basin(list(featureSource = "nwissite", featureId = gage_id))
huc08 <- nhdplusTools::get_huc(id = na.omit(unique(gdptools_weights$huc8)), type = "huc08")
huc12 <- nhdplusTools::get_huc(id = na.omit(unique(gdptools_weights$huc12)), type = "huc12")

org_par <- par(mar = c(0, 0, 0, 0))
plot(sf::st_as_sfc(sf::st_bbox(huc12)))
plot(sf::st_geometry(basin), lwd = 4, add = TRUE)
plot(sf::st_simplify(sf::st_geometry(huc08), dTolerance = 500), add = TRUE, lwd = 2)
plot(sf::st_simplify(sf::st_geometry(huc12), dTolerance = 500), add = TRUE, lwd = 0.2, border = "grey")
par(org_par)

weights <- ncdfgeom::calculate_area_intersection_weights(
  x = sf::st_transform(dplyr::select(huc12, huc12), 6931),
  y = sf::st_transform(dplyr::select(huc08, huc8), 6931),
  normalize = TRUE
)

# NOTE: normalize = TRUE means these weights can not be used for "intensive" 
# weighted sums such as population per polygon. See below for more on this.

weights <- dplyr::left_join(weights, gdptools_weights, by = c("huc8", "huc12"))

```

With weights calculated, we can do a little investigation into the differences.

```{r}

weights$diff <- weights$w - weights$gdptools_wght

# make sure nothing is way out of whack
max(weights$diff, na.rm = TRUE)

# ensure the weights generally sum as we would expect.
sum(weights$gdptools_wght, na.rm = TRUE)
sum(weights$w, na.rm = TRUE)
length(unique(na.omit(weights$huc8)))

# see how many NA values we have in each.
sum(is.na(weights$w))
sum(is.na(weights$gdptools_wght))

# look at cases where gptools has NA and ncdfgeom does not
weights[is.na(weights$gdptools_wght) & !is.na(weights$w),]

```

## Extended Example One

The following example illustrates the nuances between normalized and non-normalized area weights and shows more specifically how area weight intersection calculations can be accomplished for both intensive variables, like total population, and extensive variables, like average precipitation depth.

The set of polygons are a contrived but useful for the sake of demonstration.
```{r}

library(dplyr)
library(sf)
library(ncdfgeom)

g <- list(rbind(c(-1,-1), c(1,-1), c(1,1), c(-1,1), c(-1,-1)))

blue1 = sf::st_polygon(g) * 0.8
blue2 = blue1 + c(1, 2)
blue3 = blue1 * 1.2 + c(-1, 2)

pink1 = sf::st_polygon(g)
pink2 = pink1 + 2
pink3 = pink1 + c(-0.2, 2)
pink4 = pink1 + c(2.2, 0)

blue = sf::st_sfc(blue1,blue2,blue3)

pink = sf::st_sfc(pink1, pink2, pink3, pink4)

plot(c(blue,pink), border = NA)
plot(blue, border = "#648fff", add = TRUE)
plot(pink, border = "#dc267f", add = TRUE)

blue <- sf::st_sf(blue, data.frame(idblue = c(1, 2, 3)))
pink <- sf::st_sf(pink, data.frame(idpink = c(7, 8, 9, 10)))

text(sapply(sf::st_geometry(blue), \(x) mean(x[[1]][,1]) + 0.4),
     sapply(sf::st_geometry(blue), \(x) mean(x[[1]][,2]) + 0.3),
     blue$idblue, col = "#648fff")

text(sapply(sf::st_geometry(pink), \(x) mean(x[[1]][,1]) + 0.4),
     sapply(sf::st_geometry(pink), \(x) mean(x[[1]][,2])),
     pink$idpink, col = "#dc267f")

sf::st_agr(blue) <- sf::st_agr(pink) <- "constant"
sf::st_crs(pink) <- sf::st_crs(blue) <- sf::st_crs(5070)
```

Here, we compare blue (source) and pink (target) and w is the fraction of blue (source) that is covered pink (target) since we set `normalize = FALSE`.

```{r}

(blue_pink_norm_false <- 
calculate_area_intersection_weights(blue, pink, normalize = FALSE))
```

NOTE: `normalize = FALSE` so weights sum to 1 per _source_ polygon only when a source polygon is fully covered by the target. The non-intersecting portion is not included.

The following breaks down how to use these weights for _intensive_ or _extensive_ variables.

```{r}
blue$val = c(30, 10, 20)
blue$area_blue <- as.numeric(sf::st_area(blue))

(result <- st_drop_geometry(blue) |>
  left_join(blue_pink_norm_false, by = "idblue"))
```

For _intensive_ variables, such as precipitation depth, to calculate the value for pink-9, we calculate an area-weighted mean:

```{r}

((10 * 0.375 * 2.56) + (20 * 0.604167 * 3.6864)) / ((0.375 * 2.56) + (0.604167 * 3.6864))

```

This is saying that 0.375 of blue-3 covers pink-9 and 0.6 of blue-2 covers pink-9.
Since we are calculating a area weighted mean for an intensive variable, we 
multiply the fraction of each source polygon by its area and the value we want 
to create an area weight for. We sum the contributions from blue-2 and blue-3 
to pink-9 and divide by the sum of the combined area weights. 

NOTE: Because there is no contribution to pink-9 over some parts of the polygon, 
that missing area does not appear. The intersecting areas are 0.96 and 2.23 
meaning that we are missing 

4 - 0.96 - 2.23 = 0.81 

and could rewrite the value for pink-9 as:

```{r}
((10 * 0.375 * 2.56) + (20 * 0.604167 * 3.6864)) + (NA * 1 * 0.81) / 
  ((1 * 0.81) + (0.375 * 2.56) + (0.604167 * 3.6864))
```
Which evaluates to NA. This is why for this operation we usually drop NA terms!

For _extensive_ variables, such as population, to calculate the value of pink-9 we would do:


```{r}

((10 * 0.375) + (20 * 0.604167))

```
This is saying that 0.375 of blue-3 covers pink-9 and 0.6 of blue-2 covers pink-9.
Since we are apportioning the counts of an intensive variable from blue to
pink, the weight can be used directly without calculating an area weighted mean.

In practice, the above can be accomplished with `dplyr::group_by()` and `dplyr::summarize()`

```{r}
(result <- result |>
  group_by(idpink) |> # group so we get one row per target
  # now we calculate the value for each `pink` with fraction of the area of each
  # polygon in `blue` per polygon in `pink` with an equation like this:
  summarize(
    new_val_intensive = sum((val * w * area_blue)) / sum(w * area_blue),
    new_val_extensive = sum(val * w)))
```

Now let's do the same thing but with `normalize = TRUE`. In this case, `area` is built into `w` so you don't need to know it when using `w` for intensive variables. The downfall with normalized weights is that they can not be used directly with `extensive` variables.

Here, we set blue as source and pink as target as was done above but change normalize to `TRUE`. 

```{r}

(blue_pink_norm_true <-
calculate_area_intersection_weights(select(blue, idblue), pink, normalize = TRUE))

```

NOTE: normalize = TRUE so weights sum to 1 per _target_ polygon. Non-overlap is ignored as if it does not exist.

The following breaks down how to use these weights for _intensive_ variables.

```{r}
(result <- st_drop_geometry(blue) |>
    left_join(blue_pink_norm_true, by = "idblue"))
```

To calculate the area weighted mean value for pink-9, we would do:

```{r}
((10 * 0.24) + (20 * 0.5568)) / (0.24 + (0.5568))
```

This is saying that the portion of pink-9 that should get the value from blue-2 is 0.3 and the portion of pink-9 that should get the value from blue-3 is 0.7. In this form, our weights include the relative area of the source polygons. 

As shown above as well, the calculation can be accomplished with:

```{r}
(result <- result |>
    group_by(idpink) |> # group so we get one row per target
    # now we calculate the value for each `pink` with fraction of the area of each
    # polygon in `blue` per polygon in `pink` with an equation like this:
    summarize(
      new_val = sum((val * w)) / sum(w)))
```

## Extended Example Two

We can look at a more typical arrangement of polygons and look at this a different way. The following polygons are used below to illustrate transferring data from one set of polygons to another.

```{r, echo=FALSE}

g <- list(rbind(c(-1,-1), c(1,-1), c(1,1), c(-1,1), c(-1,-1)))

blue1 = st_polygon(g) * 0.75 + c(-.25, -.25)
blue2 = blue1 + 1.5
blue3 = blue1 + c(0, 1.5)
a4 = blue1 + c(1.5, 0)

pink1 = st_polygon(g)
pink2 = pink1 + 2
pink3 = pink1 + c(0, 2)
pink4 = pink1 + c(2, 0)

blue = st_sfc(blue1,blue2, blue3, a4)
pink = st_sfc(pink1, pink2, pink3, pink4)

pink <- st_sf(pink, data.frame(idpink = c(1, 2, 3, 4)))
blue <- st_sf(blue, data.frame(idblue = c(6, 7, 8, 9)))

plot(st_geometry(pink), border = "#dc267f", lwd = 3)
plot(st_geometry(blue), border = "#648fff", lwd = 3, add = TRUE)

text(sapply(st_geometry(blue), \(x) mean(x[[1]][,1]) + 0.4),
     sapply(st_geometry(blue), \(x) mean(x[[1]][,2]) + 0.3),
     blue$idblue, col = "#648fff")

text(sapply(st_geometry(pink), \(x) mean(x[[1]][,1]) - 0.4),
     sapply(st_geometry(pink), \(x) mean(x[[1]][,2]) - 0.5),
     pink$idpink, col = "#dc267f")

st_agr(blue) <- st_agr(pink) <- "constant"
st_crs(pink) <- st_crs(blue) <- st_crs(5070)
```

Let's also look at the values.
```{r, echo = FALSE}
blue$val <- c(1, 2, 3, 4)
blue$blue_areasqkm <- 1.5 ^ 2

plot(blue["val"], reset = FALSE, pal = heat.colors)
plot(st_geometry(pink), border = "#dc267f", lwd = 3, add = TRUE, reset = FALSE)
plot(st_geometry(blue), border = "#648fff", lwd = 3, add = TRUE)

text(sapply(st_geometry(blue), \(x) mean(x[[1]][,1]) + 0.4),
     sapply(st_geometry(blue), \(x) mean(x[[1]][,2]) + 0.3),
     blue$idblue, col = "#648fff")

text(sapply(st_geometry(pink), \(x) mean(x[[1]][,1]) - 0.4),
     sapply(st_geometry(pink), \(x) mean(x[[1]][,2]) - 0.5),
     pink$idpink, col = "#dc267f")
```

In the example code below, values from blue are translated to values from pink
using two methods: an weighted sum is done for _extensive_ data such as 
population data and a weighted mean is done for _intensive_ data such
as precipitation depth.

```{r}
# say we have data from `blue` that we want sampled to `pink`.
# this gives the percent of each `blue` that intersects each `pink`

(blue_pink <- calculate_area_intersection_weights(
  select(blue, idblue), select(pink, idpink), normalize = FALSE))

# NOTE: `w` sums to 1 per `blue` in all cases

summarize(group_by(blue_pink, idblue), w = sum(w))

# Since normalize is false, we apply weights like:
st_drop_geometry(blue) |>
  left_join(blue_pink, by = "idblue") |>
  group_by(idpink) |> # group so we get one row per `pink`
  # now we calculate the value for each `pink` with fraction of the area of each
  # polygon in `blue` per polygon in `pink` with an equation like this:
  summarize(
    new_val_intensive = sum( (val * w * blue_areasqkm) ) / sum(w * blue_areasqkm),
    new_val_extensive = sum( (val * w) ))

# NOTE: `w` is the fraction of the polygon in `blue`. We need to multiply `w` by the
# unique area of the polygon it is associated with to get the weighted mean weight.

# we can go in reverse if we had data from `pink` that we want sampled to `blue`

(pink_blue <- calculate_area_intersection_weights(
  select(pink, idpink), select(blue, idblue), normalize = FALSE))

# NOTE: `w` sums to 1 per `pink` (source) only where `pink` is fully covered by `blue` (target).

summarize(group_by(pink_blue, idpink), w = sum(w))

# Now let's look at what happens if we set normalize = TRUE. Here we
# get `blue` as source and `pink` as target but normalize the weights so
# the area of `blue` is built into `w`.

(blue_pink <- calculate_area_intersection_weights(
  select(blue, idblue), select(pink, idpink), normalize = TRUE))

# NOTE: if we summarize by `pink` (target) `w` sums to 1 only where there is full overlap.

summarize(group_by(blue_pink, idpink), w = sum(w))

# Since normalize is false, we apply weights like:
st_drop_geometry(blue) |>
  left_join(blue_pink, by = "idblue") |>
  group_by(idpink) |> # group so we get one row per `pink`
  # now we weight by the percent of each polygon in `pink` per polygon in `blue`
  summarize(new_val = sum( (val * w) ) / sum( w ))

# NOTE: `w` is the fraction of the polygon from `blue` overlapping the polygon from `pink`.
# The area of `blue` is built into the weight so we just sum the with times value per polygon.
# We can not calculate intensive weighted sums with this weight.

```

```{r cleanup, echo=FALSE}
options(org_options)
unlink("climdiv_prcp.nc")
```
